Project Description -- Erosion, Transport, Anomalous Diffusion, and Collapse in Granular and Porous Media

### 2019-11-15 docs/proposals/NSF/ProjDesc/ProjDesc.tex
added in a new interesting (I think) result

Next, we plan to extend the simulation framework to handle chemical dissolution of solid bodies via physical laws~\eqref{Ceq}--\eqref{DissVn}. For this goal, much of the computational infrastructure, such as the boundary-integral solver for the Stokes equations, is already in place, but the coupling to the advection-diffusion equations~\eqref{Ceq} and interface law~\eqref{DissVn} need to be implemented.  We proposed to solve the advection diffusion equation~\eqref{Ceq} by first using Strang splitting~\cite{str1968} to decompose the PDE into linear advection and diffusion.  The advection equation will be solved with a semi-Lagrangian method~\cite{rob1981} so that it can be solved on an Eulerian grid without a restrictive CFL condition. PI Quaife has experience with using semi-Lagrangian methods for another type of complex viscous flow~\cite{kab-qua-bir2017}.  The diffusion equation is more challenging to accurately solve in the complex porous geometries.  Two approaches include a memory-intensive space-time heat kernel~\cite{bar-eps-gre-jia-wan2019, jia-gre-wan2015, li-gre2009}, and Rothe's method coupled with a Yukawa equation solver~\cite{kro-qua2010, cau-cho-chr-sea2016}. A particular challenge of the latter method is extending functions from the complex geometry and into a regular geometry, such as a square, so that a high-order volume integral or Fourier method can be applied, and some groups have recently developed new techniques to form such an extension~\cite{fry-kro-tor2019, fry-leh-tor2019}. Alternatively, we propose to use a Laplace Transform to solve the diffusion equation
\begin{alignat}{3}
  \pd{c}{t} &= \frac{1}{\Pe} \Delta c, \quad &&\xx \in \Omega, \: t>0.
\end{alignat}
Applying the Laplace transform, $C(\xx,s) = \mathcal{L}[c](\xx,t)$ satisfies
\begin{align}
  sC(\xx,s) - \frac{1}{\Pe}\Delta C(\xx,s) = c_0(\xx), 
    \quad \xx \in \Omega,
\end{align}
where $c_0(\xx)$ is the initial concentration. This PDE will be solved with high-order accuracy and efficiently by using fast-multipole-accelerated integral equation methods of the Yukawa equation~\cite{kro-qua2011}, and the concentration will be recovered with an inverse Laplace transform~\cite{jos-war2006}.  
Once a high-order advection-diffusion solver is in place, by combining it with our integral equation formulation for the Stokes equation and the shear stress, simulations of simultaneous erosion and dissolution, while also including transport and sedimentation, would finally become possible. Such a framework could accurately simulate the geophysically relevant scenario of karst conduits with embedded granular media. The gradual changes in the medium and karst boundaries due to erosion and dissolution could create conditions favorable for mechanical failure or collapse of the structure, for example sinkholes.

@@

While the tortuosity characterizes the amount of winding of streamlines, the dispersion characterizes the spreading of the trajectories.  Dispersion can be described using the first- and second-ensemble moments of the trajectories
\begin{align}
  \langle \lambda \rangle (t) = \frac{1}{N_p} \sum_{j=1}^{N_p}
    \lambda_j(t), \quad \sigma_\lambda^{2}(t) = \frac{1}{N_p}
    \sum_{j=1}^{N_p} (\lambda_j(t) - \langle \lambda \rangle(t))^2
\end{align}
Then, the standard deviation $\sigma_\lambda$ characterizes the dispersion, and long-time behavior typically results in a power-law scaling $\sigma_\lambda \sim t^{\alpha}$. The power $\alpha$, which quantifies the long-time spreading, is $1$ in a ballistic regime which occurs at early times when the trajectories have not explored much of the porous geometry. However, once the trajectories have traversed a few grains, dispersion in porous media is often super-dispersive with $\alpha \in (0.5,1)$~\cite{dea-leb-den-tar-bol-dav2013}. In Figure~\ref{fig:100dispersion}, we investigate spreading at several different porosities of a porous geometry that is initialized with 100 grains.
\begin{wrapfigure}[]{r}{0.4\textwidth}
  \includegraphics[width=0.4\textwidth]{figs/100b_second_moment_long_ref}
  \caption{\label{fig:100dispersion} The particle spreading,
  characterized with the standard deviation $\sigma_\lambda(t)$ at 6
  different porosities. At early times, the line of slope 1 (black
  curve) indicates a ballistic motion. Then, at later times, the
  dispersion becomes super-dispersive at a rate indicated by the slope
  of the black dashed-dotted line.}
\end{wrapfigure}

@@

We have performed a carefully chosen simulation to better understand the
effect of erosion on the pore size. We start with a single grain inside
the channel that is sufficiently large that the spacing between the
initial grain and the channel walls, which is 2 units wide), is
$10^{-3}$. Since we use a Barycentric quadrature rule, we are able to
stably simulate erosion with only 512 points on the eroding body which
corresponds to an arclength spacing 12 times larger than the original
separation distance---to achieve a comparable accuracy with the spectral
trapezoid rule, 60 times as many discretization points would be
required. In Figure~\ref{fig:}

@@

\begin{wrapfigure}[]{r}{0.4\textwidth}
  \includegraphics[width=0.4\textwidth]{figs/porelets_geom} \\ \\ \\
  \includegraphics[width=0.4\textwidth]{figs/porelets_size}
\end{wrapfigure}

\todo[inline]{Talk about how to develop models for the pore sizes as a
function of time. See results from S-H}

>>>>

Next, we plan to extend the simulation framework to handle chemical dissolution of solid bodies via physical laws~\eqref{Ceq}--\eqref{DissVn}. For this goal, much of the computational infrastructure, such as the boundary-integral solver for the Stokes equations, is already in place, but the coupling to the advection-diffusion equations~\eqref{Ceq} and interface law~\eqref{DissVn} need to be implemented.  We proposed to solve the advection diffusion equation~\eqref{Ceq} by first using Strang splitting~\cite{str1968} to decompose the PDE into linear advection and diffusion.  The advection equation will be solved with a semi-Lagrangian method~\cite{rob1981} so that it can be solved on an Eulerian grid without a restrictive CFL condition. PI Quaife has experience with using semi-Lagrangian methods for another type of complex viscous flow~\cite{kab-qua-bir2017}.  The diffusion equation is more challenging to accurately solve in the complex porous geometries.  Two approaches include a memory-intensive space-time heat kernel~\cite{bar-eps-gre-jia-wan2019, jia-gre-wan2015, li-gre2009}, and Rothe's method coupled with a Yukawa equation solver~\cite{kro-qua2010, cau-cho-chr-sea2016}. A particular challenge of the latter method is extending functions from the complex geometry and into a regular geometry, such as a square, so that a high-order volume integral or Fourier method can be applied, and some groups have recently developed new techniques to form such an extension~\cite{fry-kro-tor2019, fry-leh-tor2019}. Alternatively, we propose to use a Laplace Transform to solve the diffusion equation
\begin{alignat}{3}
  \pd{c}{t} &= \frac{1}{\Pe} \Delta c, \quad &&\xx \in \Omega, \: t>0.
\end{alignat}
Applying the Laplace transform, $C(\xx,s) = \mathcal{L}[c](\xx,t)$ satisfies
\begin{align}
  sC(\xx,s) - \frac{1}{\Pe}\Delta C(\xx,s) = c_0(\xx), 
    \quad \xx \in \Omega,
  \label{eqn:DiffusionLaplace}
\end{align}
where $c_0(\xx)$ is the initial concentration. This PDE will be solved with high-order accuracy and efficiently by using fast-multipole-accelerated integral equation methods of the Yukawa equation~\cite{kro-qua2011}, and the concentration will be recovered with an inverse Laplace transform~\cite{jos-war2006}.  
Once a high-order advection-diffusion solver is in place, by combining it with our integral equation formulation for the Stokes equation and the shear stress, simulations of simultaneous erosion and dissolution, while also including transport and sedimentation, would finally become possible. Such a framework could accurately simulate the geophysically relevant scenario of karst conduits with embedded granular media. The gradual changes in the medium and karst boundaries due to erosion and dissolution could create conditions favorable for mechanical failure or collapse of the structure, for example sinkholes.

@@

While the tortuosity characterizes the amount of winding of streamlines, the dispersion characterizes the spreading of the trajectories.  Dispersion can be described using the first- and second-ensemble moments of the trajectories
\begin{align}
  \langle \lambda \rangle (t) = \frac{1}{N_p} \sum_{j=1}^{N_p}
    \lambda_j(t), \quad \sigma_\lambda^{2}(t) = \frac{1}{N_p}
    \sum_{j=1}^{N_p} (\lambda_j(t) - \langle \lambda \rangle(t))^2
\end{align}
Then, the standard deviation $\sigma_\lambda$ characterizes the dispersion, and long-time behavior typically results in a power-law scaling $\sigma_\lambda \sim t^{\alpha}$. The power $\alpha$, which quantifies the long-time spreading, is $1$ in a ballistic regime which occurs at early times when the trajectories have not explored much of the porous geometry. However, once the trajectories have traversed a few grains, dispersion in porous media is often super-dispersive with $\alpha \in (0.5,1)$~\cite{dea-leb-den-tar-bol-dav2013}. In Figure~\ref{fig:100dispersion}, we investigate spreading at several different porosities of a porous geometry that is initialized with 100 grains.
\begin{wrapfigure}[22]{r}{0.4\textwidth}
  \includegraphics[width=0.4\textwidth]{figs/100b_second_moment_long_ref}
  \caption{\label{fig:100dispersion} The particle spreading,
  characterized with the standard deviation $\sigma_\lambda(t)$ at 6
  different porosities. At early times, the line of slope 1 (black
  curve) indicates a ballistic motion. At later times, the
  dispersion becomes super-dispersive at a rate indicated by the slope
  of the black dashed-dotted line.}
\end{wrapfigure}

@@

We have performed a carefully chosen simulation to better understand the
effect of erosion on the pore size. We start with a single grain inside
the channel that is sufficiently large that the spacing between the
initial grain and the channel walls, which is 2 units wide), is
$10^{-3}$. Since we use a Barycentric quadrature rule, we are able to
stably simulate erosion with only 512 points on the eroding body which
corresponds to an arclength spacing 12 times larger than the original
separation distance---to achieve a comparable accuracy with the spectral
trapezoid rule, 60 times as many discretization points would be
required. In Figure~\ref{fig:porelets}(a), different shades of grey
correspond to the configuration of the eroding grain at five equally
spaced time steps. At three of the time steps, we plot the velocity
between the grain and the solid wall (black line). At all three
configurations, the flow is nearly parabolic indicating that  porelet
model~\cite{dea-qua-bir-jua2018} is appropriate. After performing a
thorough analysis of the pore size growth in this simple configuration,
we will develop a more complete model for networks of pores that are
prevalent in porous media.

@@

\begin{figure}[htp]
  \centering
  \begin{subfigure}[b]{0.47\textwidth}
  \begin{center}
  \includegraphics[height=0.55\textwidth]{figs/porelets_geom}
  \end{center}
  \caption{}
  \end{subfigure}
  \begin{subfigure}[b]{0.47\textwidth}
  \begin{center}
  \includegraphics[height=0.55\textwidth]{figs/porelets_size}
  \end{center}
  \caption{}
  \end{subfigure}
  \caption{\label{fig:porelets} (a) An eroding body at five equally time steps in $[0,1]$. The far field condition is a constant pressure drop in a channel three times as wide as the eroding grain.  At the three intermediate time steps, the velocity field in the field, which is nearly parabolic, is included. (b) The pore size (black) and maximum velocity in the pore (red) as a function of time.  As the channel opens, the flow rate increases to maintain a constant pressure drop, and this increases the rate of erosion.} 
\end{figure}
